---
title: "hw04-Tidy data and joins"
output:
  github_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Introduction

The goal of this homework is to solidify data wrangling skills by working some realistic problems in the grey area between data aggregation and data reshaping.

For this assignment, we have been tasked with two prompts. The first will be a data reshaping prompt while the second will be a data joining prompt.

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(knitr)
library(gapminder)
```

# Data Reshaping (Activity #2)

**Problem**: Make a tibble with one row per year and columns for life expectancy for two or more countries. 

**Solution**: Let's make a table comparing life expectancies between Canada and Italy over the years sampled in the `gapminder` dataset. Specifically, we want each row to represent a year and have a column for each country that records life expectancy in that year.
```{r}
table1 <- gapminder %>% 
  filter(country == "Canada" | country == "Italy") %>% 
  select(year,country,lifeExp) %>% 
  spread(key=country, value=lifeExp) 
kable(table1, caption = "LifeExp in Canada vs Italy",
      table.placement = "tbp", 
      caption.placement = "top", # table and title position
      col.names = c(" Year ", " LifeExp_Canada ", " LifeExp_Italy ")) # add title, edit columns names
```

From the table above, we get the lifeExp for Canada and France. Now, let's make a plot of life expectancy for Canada against France which will be visually appealing.

```{r}
table1 %>% 
  ggplot(aes(year,Canada/Italy))+
  geom_point(size=2)+
  geom_line(color="red",size=1)+
  geom_text(aes(label = year),hjust = 0.5, vjust = -0.5)+
  # adjust the position of marks
  labs(title="Life Expectancy in Canada vs Italy",x = "Year", y = "LifeExp_Canada / LifeExp_Italy")+
  # add title, lable x and y axis
  theme(plot.title = element_text(hjust = 0.5)) #center the title

```

From the graph above, it's clear that Canada has a higher life expectancy over Italy except in 1997 and 2002. And the difference between the life expectancies of these two countries is going to be smaller and smaller.


# Data Joining (Activity #1)

**Problem**: Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. 

**Solution**: Let's use `LifeCycleSavings` which is a built-in data frame in r as the second data frame. `LifeCycleSavings` is a dataset which contains 50 observations on 5 variables.

| Variable | Description |
|------------|:------------:| 
|sr| aggregate personal savings|
|pop15| % of population under 15|
|pop75| % of population over 75|
|dpi|   real per-capita disposable income|
|ddpi|  growth rate of dpi|

Let's take a look at this dataset and select two varibles from it.The new dataset we get is called `mytable`.
```{r}
mytable <- LifeCycleSavings %>% 
  select(sr,dpi) %>% 
  rownames_to_column("country") %>% 
# convert the row names into first column
  mutate_if(is.factor, as.character)

kable(mytable)
```

## mutating join


### left_join and right_join
Let's use left_join and right_join seperately to join part of gapminder(4 columns, 142 rows) and mytable(3 columns, 50 rows) together.
```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>%  #add a restriction to get a smaller data set %>% 
  left_join(mytable,by=c("country")) %>% 
  head() %>% # the table is too long so I just present some rows here
  kable()
```

```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>% #add a restriction to get a smaller data set
  right_join(mytable,by=c("country")) %>% 
  head() %>% # the table is too long so I just present some rows here
  kable()
```

* `left_join(x,y)` return all rows from x. Join matching rows from `y` to `x`. Rows in x with no match in y will show NA in the new columns.
As we only have 50 rows in `mytable` and while 142 rows in the selected part of gapminder, NA is displayed to represent some unavailable data. 
The final table contains 142 rows.

* `right_join(x,y)` return all rows from y. Join matching rows from `x` to `y`. Rows in y with no match in x will show NA in the new columns. 
Similarly, we get some NA due to lack of information in the selected part of gapminder.
The final table contains 50 rows.

### inner_join
If we only want those completed rows, we can drop those with NA by inner_join.
```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>% #add a restriction to get a smaller data set
  inner_join(mytable,by=c("country")) %>% 
  head() %>% # the table is too long so I just present some rows here
  kable()
```

* `inner_join(x,y)` retain only rows in both sets.
The final table contains 45 rows.

### full_join
If we want all the rows from both data set, we can use full_join.
```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>% #add a restriction to get a smaller data set
  full_join(mytable,by=c("country")) %>% 
  head() %>% # the table is too long so I just present some rows here
  kable()
```

* `full_join(x,y)` return all rows and columns from both x and y. For the unmatching values,it returns NA for the missing one.
The final table contains 147 rows.

## filtering join

### semi_join
If we only want the data from the gapminder which matches data from mytable, semi_join is a good choice which not show any information from mytable. 
```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>% #add a restriction to get a smaller data set
  semi_join(mytable,by=c("country")) %>% 
  head() %>% 
  kable()

```

* `semi_join(x,y)`return only columns from x which have matching values in y. A semi join differs from an inner join because an inner join returns the columns from both tables, while an semi join only return columns from x.
The final table contains 45 rows and 4 columns while the table resulted from inner join contains 45 rows and 6 columns.

### anti_join
In some cases, we want to see all the data from the part of gapminder which do not have a match in mytable, anti join is a good choice. 
```{r}
gapminder %>% 
  filter(year == 1967) %>% 
  select(country,continent,year,gdpPercap) %>% #add a restriction to get a smaller data set
  anti_join(mytable,by=c("country")) %>% 
  head() %>% 
  kable()

```

* `anti_join(x,y)` return all rows from x which do not have matching values in y.
The final table contains 97 rows.

